---
title: "Gobi: Clustering"
author: "Marie Hackenberg"
date: "2025-02-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Install and load libraries
```{r}
library(readr)
library(Mfuzz)
library(tidyverse)
library(Biobase)
library(nparLD)
library(tidyr)
library(dplyr)
library(ggplot2)
library(readxl)
```

```{r}
sig_89_metabolites_path <- "/Users/mariehackenberg/Documents/Bioinformatik/9. Semester (WS24:25)/Gobi/Data Sheet 2.xlsx" #Marie

sig_89_metabolites <- read_xlsx(sig_89_metabolites_path, "TABLE S4 clean", col_names = TRUE)

```


# Load data: INPUT MUSS AUCHT T-TEST TREATED SEIN
```{r}
#cl_data_path <- "../output/preproc_z_merged_fl_1.csv"
cl_data_path <- "../output/pp_imp_merged.csv"
sig_p_for_3_path <- "/Users/mariehackenberg/Documents/Bioinformatik/9. Semester (WS24:25)/Gobi/Data Sheet 2.xlsx" #Marie
cl_data <- read.csv(cl_data_path, sep = ",", header = TRUE)

sig_p_for_3 <- read_xlsx(sig_p_for_3_path, "TABLE S2", col_names = TRUE)
colnames(sig_p_for_3) <- sig_p_for_3[2, ]
sig_p_for_3 <- cbind(sig_p_for_3[, 1], sig_p_for_3[, ncol(sig_p_for_3) - 1])
sig_p_for_3[is.na(sig_p_for_3)] <- 0
sig_met_for_3 <- sig_p_for_3 %>%
  filter(`sig for 3 challenges` == "TRUE")

core_pp_met <- read_xlsx(sig_p_for_3_path, "TABLE S4", col_names = FALSE)
colnames(core_pp_met) <- core_pp_met[4, ]
core_pp_met <- core_pp_met[5:nrow(core_pp_met), 1]

any(core_pp_met$metabolite %in% colnames(cl_data))
all(core_pp_met$metabolite %in% colnames(cl_data))
missing_core_metabolites <- core_pp_met$metabolite[!core_pp_met$metabolite %in% colnames(cl_data)]
missing_core_metabolites <- rep(sort(missing_core_metabolites), length.out = nrow(missing_core_metabolites))
print(missing_core_metabolites)

#TEMP
core_pp_met$metabolite <- rep(sort(core_pp_met$metabolite), length.out = nrow(core_pp_met))
met <- colnames(cl_data[, -c(1:4)])
cl_data_metab <- data.frame(cl_data_metab = colnames(cl_data[, -c(1:4)]))
new_object$cl_data_metab <- rep(sort(new_object$cl_data_metab), length.out = nrow(new_object))
new_object$core_pp_met <- c(core_pp_met$metabolite, rep("_", length.out = nrow(new_object) - length(core_pp_met$metabolite)))

common_values <- intersect(new_object$cl_data_metab, new_object$core_pp_met)
new_object_filtered <- new_object %>%
  filter(!cl_data_metab %in% common_values & !core_pp_met %in% common_values)
#new_object$core_pp_met <- new_object %>%
 # filter(!cl_data_metab %in% common_values & !core_pp_met %in% common_values)

#new_object$core_pp_met <- core_pp_met$metabolite



## data[is.na(data)] <- 0
# data <- data %>%
#   mutate(challenge = case_when(
#     time >= 1 & time <= 9 ~ "fasting",      # time 1-10 corresponds to fasting
#     time >= 33 & time <= 39 ~ "exercise",    # time 33-39 corresponds to exercise
#     time >= 40 & time <= 49 ~ "oltt",        # time 40-48 corresponds to OLTT
#     TRUE ~ NA_character_                                            # assign NA for any time outside ranges
#     )) %>%
#   filter(!is.na(challenge)) %>%  # remove rows where 'challenge' is NA
#   dplyr::select(challenge, everything()) # make challenge the first column
```
# Fuzzy c-means clustering

## Prepare data for Mfuzz
```{r}
# SKIP ONLY FOR pp_imp

# # Convert value types
# options(digits = 14) # set decimal precision to original precision in raw data
# convert_to_numeric <- function(x) {
#   as.numeric(gsub(",", ".", x)) # R uses . for decimal numbers
# }
# 
# cl_data <- cl_data %>%
#   mutate(across(
#     -c(subject, time, challenge), # exclude subject, challenge_time, and challenge
#     convert_to_numeric # apply conversion function
#   ))

# Convert challenge time to categorical
#data$challenge_time <- as.factor(data$challenge_time)
```
## Normalize metabolite data already done in downloaded dataset
```{r}
# SKIP
# # Z-score transformation
# z_score <- function(x){
#   (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)
# }
# 
# # Apply z-score function to all metabolite columns

# data[metabolite_columns] <- data[metabolite_columns] %>% apply(2, z_score)

```

## Reshape data for Mfuzz
rows = metabolites, columns = time points (concatenated across subjects & challenges)
```{r}
# FOR PP_IMP! #todo: also for zscore?
cl_data <- cl_data %>%
  dplyr::select(-platform_name.x)
cl_data <- cbind(platform_name = cl_data$platform_name,
                 cl_data %>%
                   dplyr::select(-platform_name) %>%
                   dplyr::select(-platform_name.y))

metabolite_columns <- names(cl_data)[5:ncol(cl_data)] # 5 for pp_imp. #todo: 4 for zscore?

# Convert challenge_time to numeric
cl_data <- cl_data %>% mutate(time = as.numeric(time))

ver <- cl_data %>%
  count(subject, challenge, time) %>%
  print(n = 20)

cl_data_long <- cl_data %>%
  pivot_longer(
    cols = all_of(metabolite_columns),  # Convert metabolite columns into rows
    names_to = "metabolite",
    values_to = "value"
  )

# Pivot data to wide format
cl_data_wide <- cl_data_long %>%
  pivot_wider( # remove non-numeric columns by itself
    names_from = c(subject, challenge, time),
    values_from = value,
    names_sep = "_"
    )
dim(cl_data_wide)
```
```{r}
# Convert to matrix
cl_data_matrix <- as.matrix(cl_data_wide[, -c(1:2)])
rownames(cl_data_matrix) <- cl_data_wide$metabolite  # assign metabolite names

eset <- new("ExpressionSet", exprs = cl_data_matrix)
```
## Estimate fuzzification parameter (m)
```{r}
m <- mestimate(eset)
print(m)
```
## Perform fuzzy c-means clustering
```{r}
set.seed(123)  # for reproducibility
clusters <- mfuzz(eset, c = 8, m = m)  # "c" is number of clusters (experiment with different values) #todo
```
## Visualise clustering results
```{r}
#plot(clusters) doesn't work so plot manually

membership <- clusters$membership

# Plot the cluster membership
matplot(t(membership), type = "l", lty = 1, col = 1:ncol(membership),
        xlab = "Metabolites", ylab = "Membership", main = "Cluster Membership")
legend("topright", legend = paste("Cluster", 1:ncol(membership)), col = 1:ncol(membership), lty = 1)

```
## Inspect cluster assignments
```{r}
# Get cluster assignments for each metabolite
cluster_results <- data.frame(metabolite = rownames(cl_data_matrix), cluster = clusters$cluster)

# Check number of metabolites in each cluster
table(cluster_results$cluster)

```

## Visualize temporal profiles for each cluster
```{r}
# Plot temporal profiles of metabolites in each cluster
for (cluster_num in 1:max(cluster_results$cluster)) {
  metabolites_in_cluster <- rownames(cl_data_matrix)[cluster_results$cluster == cluster_num]
  cluster_data <- cl_data_matrix[metabolites_in_cluster, ]
  
  # Plot the temporal profiles for the metabolites in the current cluster
  matplot(t(cluster_data), type = "l", col = 1:nrow(cluster_data), lty = 1, 
          xlab = "Time", ylab = "Metabolite Level", 
          main = paste("Temporal Profiles of Cluster", cluster_num))
}

```

```{r}
# Save cluster results to a CSV file
write.csv(cluster_results, "../output/metabolite_clusters.csv", row.names = FALSE)
```

```{r}
```
